Connectome inferences using partial correlation from imaging calcium signal

### 2014-06-13 article.tex
averaging + stacking values

\begin{abstract}
Understanding how the brain works is a key
to treat brain disorders such as Parkinson's disease,
epilepsy. Retrieving connectomes, the neurons wiring map, will shed a new
light on the anatomical and functional connectivity of the brain. In the
context of the Connectomics challenge, we propose a simple  algorithm made of
four stages, using preprocessing and partial correlation to achieve a network
inference of high quality. We put in perspective our method
with other inference methods such as GTE (\cite{stetter2012model}) and Genie3
(\cite{huynhthu2010inferring}).

@@

% BG : brain, complex organ
The human brain is a complex biological organ, formed by 100
billions of neurons with 7000 synaptic connections on average.
Cutting edge optical devices have been developed to track in real-time
the neuronal activity of thousands of neurons through a fluorescent
calcium indicator  [todo cite]. Automatic and analytical methods are required to infer
the connectomes from those time series.

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
The connectomes is represented by a (un)directed graph $G = (V, E)$ comprising
a set of nodes $V$ and a set of edges $E \subseteq \left\{(i, j) \in V \times
V\right\}$, (un)ordered pair from $V$.
Each node $i \in V$ represents a neuron, and each edge $(i, j) \in E$
represents a direct link from neuron $i$ to neuron $j$.
The network inference task is set as follows :
\textit{Given the observations $(x_t \in \mathcal{R}^{|V|})_{t=1}^T $
along $T$ periods of time of the $|V|$ neurons, the goal is to infer the
adjacency matrix $A : a_{i,j} = I((i, j) \in E)$ of the graph $G$,
where $I$ is the indicator function.}

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation\footnote{Source code is available at
\url{https://github.com/asutera/kaggle-connectomics}.} to infer the neural
network, which is the winning method of the Connectomics challenge
\footnote{\url{http://connectomics.chalearn.org}}.

@@

% Defects of the imaging device
and will have to cope with
a slower sampling rate than the neuron firing speed, a superposition
of the activity of several neurons and the slow decay of fluorescent calcium
indicator.

@@

% The main challenge in the inference of the network $y$
% is to distinguish direct from indirect interactions \cite{de2004discovery}.
% For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
% we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
% from the indirect interaction $A \rightarrow C$.

@@

\begin{table}[H]
\centering
\caption{Inverse correlation (ROC). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.777 & 0.767 & 0.772 & 0.774 \\
PCA                 & 0.780 & 0.770 & 0.776 & 0.777 \\
Weights             & 0.780 & 0.769 & 0.776 & 0.777 \\
Weights + PCA       & 0.783 & 0.772 & 0.779 & 0.780 \\
H-T                 & 0.893 & 0.891 & 0.891 & 0.886 \\
H-T + PCA           & 0.894 & 0.892 & 0.891 & 0.886 \\
H-T + Weights       & 0.899 & 0.897 & 0.896 & 0.891 \\
H-T + Weights + PCA & 0.900 & 0.898 & 0.896 & 0.892 \\
P-P                 & 0.925 & 0.925 & 0.924 & 0.923 \\
P-P + PCA           & 0.926 & 0.926 & 0.925 & 0.923 \\
P-P + Weights       & 0.931 & 0.930 & 0.928 & 0.927 \\
P-P + Weights + PCA & 0.932 & 0.931 & 0.930 & 0.928 \\
Averaging           & & & & \\
Stacking            & & & & \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Inverse correlation (P-R). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.070 & 0.064 & 0.068 & 0.072\\
PCA                 & 0.076 & 0.070 & 0.075 & 0.079\\
Weights             & 0.074 & 0.067 & 0.072 & 0.076\\
Weights + PCA       & 0.080 & 0.073 & 0.079 & 0.083\\
H-T                 & 0.264 & 0.260 & 0.269 & 0.241\\
H-T + PCA           & 0.266 & 0.263 & 0.273 & 0.244\\
H-T + Weights       & 0.280 & 0.273 & 0.281 & 0.251\\
H-T + Weights + PCA & 0.284 & 0.278 & 0.285 & 0.255\\
P-P                 & 0.322 & 0.324 & 0.323 & 0.312\\
P-P + PCA           & 0.347 & 0.352 & 0.355 & 0.341\\
P-P + Weights       & 0.333 & 0.334 & 0.327 & 0.313\\
P-P + Weights + PCA & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & & & & \\
Stacking            & & & & \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Correlation (ROC). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.681 & 0.699 & 0.683 & 0.681\\
Weights             & 0.695 & 0.713 & 0.697 & 0.694\\
H-T                 & 0.876 & 0.873 & 0.877 & 0.869\\
H-T + Weights       & 0.886 & 0.884 & 0.891 & 0.877\\
P-P                 & 0.857 & 0.850 & 0.843 & 0.832\\
P-P + Weights       & 0.826 & 0.823 & 0.817 & 0.799\\
Averaging           & & & & \\
Stacking            & & & & \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Correlation (P-R). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.028 & 0.028 & 0.025 & 0.026\\
Weights             & 0.031 & 0.030 & 0.027 & 0.028\\
H-T                 & 0.143 & 0.128 & 0.150 & 0.129\\
H-T + Weights       & 0.153 & 0.145 & 0.170 & 0.132\\
P-P                 & 0.100 & 0.087 & 0.079 & 0.083\\
P-P + Weights       & 0.079 & 0.071 & 0.067 & 0.064\\
Averaging           & & & & \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{abstract} Understanding how the brain works is a key to understand and
to treat brain pathologies and disorders such as Parkinson's disease,
epilepsy. Retrieving the connectomes, neurons connectivity map, will shed new
lights on the anatomical and functional connectivity of the brain. In the
context of the Connectomics challenge, we propose a simple  algorithm made of
four stages, using preprocessing and partial correlation to achieve a network
inference of high quality. We put in perspective our method by comparing
with other inference method such as GTE \cite{stetter2012model}, another
method proposed to infer a neural network from calcium fluorescent imaging,
and Genie3 [], a tree-based approach very competitive in several gene
inference challenge (i.e. DREAM4 and DREAM5).

@@

% BG : brain, complex organ
The human brain is a complex biological organ, which is formed by 100
billions of neurons with 7000 synaptic connections on average. Inferring the
neuron connectivity network through dissection would be an impossible
and a daunting task. Cutting edge optical devices [todo cite] allow to track
simultaneously the neural activity of thousands of neurons through a fluorescent
calcium indicator.

From those observational data, the neural connectivity inference consists in
highlighting direct relationships between neurons. However, those imaging
systems have a sampling rate smaller than the neuron firing speed and are
limited to two dimensional image, while the neuronal network is a
three dimensional structure, which may lead to superpose the observed activity
of different neurons. Furthermore, the decay of the fluorescent calcium signal
is slower than its rise.

% Introduce notation + goal in mathematical score
% TODO check convention for y_{ij}
One way to emphasize the connections between neurons is to represent them by a
(un)directed graph with $p$ nodes: each node represents a neuron, and each
edge $y_{ij}$ represents a direct link from neuron  $i$ to neuron $j$. If the
graph is undirected, then $y_{ij}$ is equal to $y_{ji}$  for all $i,j$.
The unsupervised task may be formulated as:
\textit{Given $X \in \mathcal{R}^{n \times p}$  a set of $p$ time series of
fluorescent calcium concentration of size $n$ from $p$ neurons, the goal is to
infer the connectivity graph $y \in \left\{0, 1\right\}^{p \times p}$.}
The main challenge in the inference of the network $y$
is to distinguish direct from indirect interactions \cite{de2004discovery}.
For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
from the indirect interaction $A \rightarrow C$.

\textcolor{red}{AJ: Should we keep the following paragraph? While interesting
                this might not be relevant for a 6 page abstract.}
Since only observational data are available, the network recovery is set as an
unsupervised learning problem. Whenever a partial knowledge of the target
network or known networks highly similar to the unknown one is available,
supervised learning [todo cite] and transfer learning  [todo cite] based
approach could be considered to infer networks.

% Solution
In this paper, we describe a simple and a theoretically grounded approach
based on partial correlation to infer the neural connectivity network.
This is also the winning method of the Connectomics challenge
\footnote{\url{http://connectomics.chalearn.org/}}. For reproducibility, the
source code of the method is available at
\url{https://github.com/asutera/kaggle-connectomics}  BSD 3-clauses
license.

@@

@@

@@

\begin{table}[H]
\centering
\caption{Inverse correlation (ROC). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.777 & 0.767 & 0.772 & 0.774 \\
PCA                 & 0.780 & 0.770 & 0.776 & 0.777 \\
Weights             & 0.780 & 0.769 & 0.776 & 0.777 \\
Weights + PCA       & 0.783 & 0.772 & 0.779 & 0.780 \\
H-T                 & 0.893 & 0.891 & 0.891 & 0.886 \\
H-T + PCA           & 0.894 & 0.892 & 0.891 & 0.886 \\
H-T + Weights       & 0.899 & 0.897 & 0.896 & 0.891 \\
H-T + Weights + PCA & 0.900 & 0.898 & 0.896 & 0.892 \\
P-P                 & 0.925 & 0.925 & 0.924 & 0.923 \\
P-P + PCA           & 0.926 & 0.926 & 0.925 & 0.923 \\
P-P + Weights       & 0.931 & 0.930 & 0.928 & 0.927 \\
P-P + Weights + PCA & 0.932 & 0.931 & 0.930 & 0.928 \\
Averaging (PCA)     & 0.943 & 0.942 & 0.942 & 0.939 \\
Stacking            & 0.944 & 0.943 & 0.942 & 0.940 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Inverse correlation (P-R). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.070 & 0.064 & 0.068 & 0.072\\
PCA                 & 0.076 & 0.070 & 0.075 & 0.079\\
Weights             & 0.074 & 0.067 & 0.072 & 0.076\\
Weights + PCA       & 0.080 & 0.073 & 0.079 & 0.083\\
H-T                 & 0.264 & 0.260 & 0.269 & 0.241\\
H-T + PCA           & 0.266 & 0.263 & 0.273 & 0.244\\
H-T + Weights       & 0.280 & 0.273 & 0.281 & 0.251\\
H-T + Weights + PCA & 0.284 & 0.278 & 0.285 & 0.255\\
P-P                 & 0.322 & 0.324 & 0.323 & 0.312\\
P-P + PCA           & 0.347 & 0.352 & 0.355 & 0.341\\
P-P + Weights       & 0.333 & 0.334 & 0.327 & 0.313\\
P-P + Weights + PCA & 0.364 & 0.366 & 0.359 & 0.344\\
Averaging           & 0.403 & 0.404 & 0.398 & 0.388\\
Stacking            & 0.404 & 0.405 & 0.399 & 0.389\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Correlation (ROC). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.681 & 0.699 & 0.683 & 0.681\\
Weights             & 0.695 & 0.713 & 0.697 & 0.694\\
H-T                 & 0.876 & 0.873 & 0.877 & 0.869\\
H-T + Weights       & 0.886 & 0.884 & 0.891 & 0.877\\
P-P                 & 0.857 & 0.850 & 0.843 & 0.832\\
P-P + Weights       & 0.826 & 0.823 & 0.817 & 0.799\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Correlation (P-R). Improvement of each stage of our approach. Note that we choose the
         \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
\begin{tabular}{*{5}{l}}
\toprule
Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
\midrule
Nothing             & 0.028 & 0.028 & 0.025 & 0.026\\
Weights             & 0.031 & 0.030 & 0.027 & 0.028\\
H-T                 & 0.143 & 0.128 & 0.150 & 0.129\\
H-T + Weights       & 0.153 & 0.145 & 0.170 & 0.132\\
P-P                 & 0.100 & 0.087 & 0.079 & 0.083\\
P-P + Weights       & 0.079 & 0.071 & 0.067 & 0.064\\
\bottomrule
\end{tabular}
\end{table}
